Grappa: Scalable Computing on Mass-Market Hardware for Irregular, Low-Locality Applications

### 2012-07-22 doc/asplos13/implementation.tex
Hacked out and simplified grappa and restructured implementation

\subsubsection{Tasks and workers}

Grappa tasks are 32-byte entities: a 64-bit function pointer plus
three 64-bit arguments. We use three arguments because tasks are
usually generated as part of a parallel loop decomposition, and thus
each task needs three kinds of data.
\begin{description}
\item[Function pointer] This 64-bit value indicates what routine
  should run. We configure the system nodes to disable address randomization, so that we can depend on the fact that function pointers are valid across process images.
\item[Private argument] This 64-bit value is typically used for task-specific
  data (e.g., a loop index).
\item[Shared argument] This 64-bit value is typically used for data shared
  between all tasks that are part of a loop or to specify the number
  of loop iterations.
\item[Synchronization argument] This 64-bit value is typically used to determine
  when all tasks that are part of a loop have finished. It is usually
  a global pointer to a synchronization object allocated at the core
  that spawned the task.
\end{description}
While these are the most common uses of the three task arguments, they
are treated as arbitrary 64-bit values in the runtime, and can be used
for any purpose.

@@

\subsubsection{Context switching}

Context switching in Grappa is non-preemptive. As with other cooperative
multithreading systems, we treat context switches as function calls,
saving and restoring only the callee-saved state as specified in the
x86-64 ABI \cite{someone}. This involves saving six general-purpose
64-bit registers and the stack pointer, as well as the 16-bit x87
floating point control word and the SSE context/status register. Thus,
the minimum amount of state a cooperative context switch routine must
save according to the ABI is 62 bytes.

Since the compiler sees all calls to the context switch routine, we
can save even less state. Our context switch routine appears to the
compiler as inline assembly; we declare all the registers we need
to save as ``clobbered'' by the inline assembly routine, and the
compiler will issue its own save and restore code as needed. This allows the
compiler to avoid saving any registers that are not used, or are used
for temporary values that are not needed after the context switch.
\comment{I highly doubt this is important in real apps beyond our kernels.  -M
Why not? liveness is a common compiler analysis -BM
Because it has to be live across the entire application.  All real applications do some form of floating point, however minor... -M

In
particular, only one of our applications does floating point
operations; all the other applications can safely ignore the x87 and
SSE2 registers as part of context switching.}

\subsubsection{Scheduling}

Each core in the Grappa system has its own independent scheduler. Each
scheduler has three main tasks to perform.  First, it must ensure that
its communication resources are serviced reasonably often. Second, it
must ensure that if a running task was waiting on a long-latency
operation and that operation has completed, the task will be
rescheduled. Third, if there are tasks waiting to be run and spare
execution resources, the tasks must be matched with idle workers.

@@

@@

\subsubsection{Work stealing} \label{subsec:implementation-worksteal}

When a \checkme{grappa-process} runs out of work, it becomes a ``thief''
and asks another other \checkme{grappa-process} (``victim'') for work.
If the victim has tasks to spare in its public queue, they will be
transferred to the thief's public queue.

\comment{
where to put the following related? -BM

Fork join computations: implement with futures or continuation passing
style
feed-forward computations: implement with scalable task counting (we
focus on this style in our benchmarks)
(briefly describe global task joiner)

choices we've made
done -random victims
hmm..  -steal from FIFO side of queue
hmmm... -steal when at the moment readyQ empty and task queues empty
done -steal min(chunk\_size, half)

omit - Recent work has explored scalable work stealing in terms of efficient
termination detection. 
}

\subsubsection{Network Aggregation}

@@

>>>>

\paragraph{Tasks and workers} Grappa tasks are 32-byte entities: a
64-bit function pointer plus three 64-bit arguments. We use three
arguments because tasks are usually generated as part of a parallel loop
decomposition, and thus each task needs three kinds of data.
\begin{description}
\item[Function pointer] This 64-bit value indicates what routine
  should run. We configure the system nodes to disable address randomization, so that we can depend on the fact that function pointers are valid across process images.
\item[Private argument] This 64-bit value is typically used for task-specific
  data (e.g., a loop index).
\item[Shared argument] This 64-bit value is typically used for data shared
  between all tasks that are part of a loop or to specify the number
  of loop iterations.
\item[Synchronization argument] This 64-bit value is typically used to determine
  when all tasks that are part of a loop have finished. It is usually
  a global pointer to a synchronization object allocated at the core
  that spawned the task.
\end{description}
While these are the most common uses of the three task arguments, they
are treated as arbitrary 64-bit values in the runtime, and can be used
for any purpose.

@@

\paragraph{Context switching} Grappa context switches between tasks
non-preemptively. As with other cooperative multithreading systems, we
treat context switches as function calls, saving and restoring only the
callee-saved state as specified in the x86-64 ABI \cite{someone}. This
involves saving six general-purpose 64-bit registers and the stack
pointer, as well as the 16-bit x87 floating point control word and the
SSE context/status register. Thus, the minimum amount of state a
cooperative context switch routine must save according to the ABI is 62
bytes.

Since the compiler sees all calls to the context switch routine, we
can save even less state. Our context switch routine appears to the
compiler as inline assembly; we declare all the registers we need
to save as ``clobbered'' by the inline assembly routine, and the
compiler will issue its own save and restore code as needed. This allows the
compiler to avoid saving any registers that are not used, or are used
for temporary values that are not needed after the context switch.

\paragraph{Scheduling} Each core in the Grappa system has its own
independent scheduler. Each scheduler has three main tasks to perform. 
First, it must ensure that its communication resources are serviced
reasonably often. Second, it must ensure that if a running task was
waiting on a long-latency operation and that operation has completed,
the task will be rescheduled. Third, if there are tasks waiting to be
run and spare execution resources, the tasks must be matched with idle
workers.

@@

\paragraph{Work stealing} When a Grappa node runs out of work, it
becomes a ``thief'' and asks another other Grappa nodes (``victim'') for
work. If the victim has tasks to spare in its public queue, they will be
transferred to the thief's public queue.

@@

\subsection{Network Aggregation}

@@

\comment{

Text that was in section 3 - Grappa.  Seems entirely redundant.  Not adding this anywhere.  -Mark

Grappa contains two components that are not directly visible to
the programmer but are key to providing good performance. Both depend
on Grappa's ability to tolerate latency in order to obtain this
performance.

The first is a distributed dynamic load balancing mechanism that uses
work stealing to move tasks from busy cores to idle cores. When a core
detects that it has idle execution resources and no new tasks to run, it
issues steal requests over the network. The core continues to execute
remaining active work while the steal requests are in flight. When a
steal request succeeds, it returns with more tasks that can be matched
with execution resources and scheduled.

The second is a network packet scheduling and aggregation system. From
the perspective of the application, all communication between tasks in
Grappa takes place either through global shared memory or remote task
spawns. Both of these requests are ultimately translated into explicit
inter-system-node/core communication requests that are dispatched to
the \gasnet~library. In our benchmarks, we found that most of our
packets were small (44 bytes on average for BFS). Commodity network
interfaces do not transport small packets efficiently (see
Section~\ref{sec:bandwidth-versus-packet-size}): the rate at which the
network interface will accept new packets of any size is limited, so
the only way to use a respectable fraction of the bandwidth is to send
larger messages.

Grappa's latency tolerance allows us to leverage the parallelism in
our problem domain to build larger packets. Tasks can generate
requests faster than the network will accept new packets, so instead
of immediately sending every message, Grappa queues messages heading
for the same destination. Once enough messages are queued, Grappa
bundles all the messages into a single packet and sends it to the
destination core. Tasks that require a response can suspend, allowing
their core to do other work while waiting for a reply.

While queuing adds latency to each individual request, it is a much
lighter-weight operation than sending a packet and allows us to
amoritize the cost of sending the packet across multiple tasks, while
also making better use of available network bandwidth. In this way,
Grappa trades per-request latency for application-level
throughput.
}
